---
title: "Example anisotropy analysis using Kdirectional"
author: "Tuomas Rajala"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 2D anisotropy analysis using Kdirectional}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
The package `Kdirectional` is a collection of tools for analysing the isotropy of 2- and 3-dimensional point patterns. This document shows how to use some of these tools by analysing an synthetic example pattern.

In this vignette we use the following packages:
```{r, include=FALSE}
library(spatstat)
library(Kdirectional)
library(ellipsoid)
library(sphere)
```

```{r, eval=FALSE}
library(spatstat)
library(Kdirectional)
library(ellipsoid)
library(sphere)
```



## Example pattern
We simulate Strauss process and transform it. Assume the transformation is 

```{r}
comp <- 0.7
D <- diag( c(comp, 1/comp))
angle <- pi/8
R <- sphere::rotationMatrix(az=angle)[-3,-3]
C <- R%*%D
```

that is, we compress the x-axis by factor 0.7 and stretch the y-axis to keep volume 1. Then we simulate the original process in a window that will be square after the transformation

```{r, warning=FALSE}
set.seed(1)
x0 <- rStrauss(500, gamma = 0.05, R = 0.03, W = affine(square(), solve(C)))
```

and transform it to the observed pattern
```{r}
x <- affine(x0, C)
```

Check things went ok (define a little helper for plotting):

```{r, fig.width=8}
plotp <- function(...) plot(..., cex=.3)
par(mfrow=c(1,2), mar=c(1,1,1,1))
plotp(x0)
plotp(x)
```

The pattern on the left is what we want to estimate.


## Nearest neighbour vectors
Check the distribution of nearest neighbours vectors:
```{r, fig.height=5, fig.width=5}
nna <- nnangle(coords(x))
ang <- nna[,1]-pi
angles.flower(ang, ci = T, k=12)
```

The blue rings give approximate 95% pointwise confidence intervals. Some structure seems to be visible. Nearest neighbours are however not very informative, lets analyse the full set of pairwise vectors.

## Fry-points and ellipsoids
As in the paper [1], the Fry-points tell us interesting things about the isotropy. 

```{r}
f <- fry_ellipsoids(x, nvec=1:20, nangles = 30, eps=0.1, r_adjust = .2)
```

`r_adjust=.2` reduces the computations to pairs with length within 20% window width, `eps` adds a bit of angular smoothing (in radians), and `nangles` chooses the amount of directions to look at. 

Plot the estimated ellipsoids:

```{r, fig.height=5, fig.width=5}
plot(f, zoom=.3, used_points = FALSE)
```

Oval shape is visible, some ambiguity of rotation. Check the ellipticity with a contrast $axis_x-axis_y=0$:

```{r, fig.width=5, fig.height=5}
ci <- confint(f, nsim=500)
print(summary(ci)[1:5,], digits=2)
plot(ci, ylim = c(-1,1)*.1)
```

The first contour is poorly estimated as there are some short range pairs, kind of like Poisson noise. Others seem to hint of compression (the p-values and confidence intervals are biased and conservative). 

Let's compute the average rotation, where we scale each ellipsoid to have det = 1 by setting `just_rotation=TRUE`.

```{r}
contours <- f$ellipsoids[-1]
mean_e <- mean_ellipsoids(contours, just_rotation = TRUE)
summary(mean_e)
```
Rotation seems to be non-zero (or non $\pm\pi$). Ignore the error variance, its not correct. 

```{r}
c(true=angle, est=mean_e$rot_angle + pi)
```

Backrotate the pattern:

```{r}
Rhat <- sphere::rotationMatrix(az=pi)[-3,-3]%*%mean_e$rot 
xr <- affine(x, solve(Rhat))
```

Next we estimate the compression by minimizing the difference of K-functions along the main axes. We need to provide a grid to optimise over:

```{r}
gamma_vec <- seq(0.5, 1, by=0.05)
grid <- cbind(gamma_vec, 1/gamma_vec)
```

Then compute the anisotropy profile:

```{r}
r <- seq(0, 0.2, length=50)
eps <- pi/10
ani <- anisotropy_profile_fast(xr, grid = grid, r=r, eps=eps, power=2)
ani0 <- anisotropy_profile_fast(x, grid = grid, r=r, eps=eps, power=2)
```

We set the sector-K angle with `eps` and use integral of squared differences (`power=2`) as the distance.

Plot the profiles, scaled to 0-1 for plotting (it's the default):

```{r, fig.width=6, fig.height=6}
plot(ani, type="b", scale=TRUE)
lines(ani0, type="b", col=2)
```

```{r}
( s <- summary(ani) )
```

Estimated compression

```{r}
c(true=comp, est=s$estimate[1])
```


Backtransform:

```{r, fig.width=5, fig.height=3}
Dhat <- s$estimate
xhat <- affine(xr, solve(Dhat))
```

Recap of the steps:

```{r, fig.width=6, fig.height=5}
par(mfrow=c(2,2), mar=c(1,1,1,1))
plotp(x)
plotp(xr)
plotp(xhat)
plotp(x0)
```

Finally, let's illustrate the estimated transformation. Use simulation and estimation to get an ellipsoid object:

```{r}
Chat <- Rhat %*% Dhat
e0 <- ellipsoid_OLS( rellipsoid(1000, c(1,1), R = C) )
e <- ellipsoid_OLS( rellipsoid(1000, c(1,1), R = Chat) )
```

```{r, fig.height=5, fig.width=5}
plot(e0, add=FALSE)
plot(e, col=3)
```



### References

[1] T Rajala, A Särkkä, C Redenbach, M Sormani (2015): Estimating geometric anisotropy in spatial point patterns, Spatial Statistics







